Evaluating infectious disease forecasts with allocation scoring rules

### 2023-07-17 alloscore_manuscript/allocation.tex
add some uncommitted commented text from another branch, related to review of forecast eval via decision theory

Here we pursue one possible tether between multivariate forecasts and policy: the allocation of limited medical resources in response to COVID-19 hospitalizations in various regions so as to minimize expected unmet need. Given probabilistic forecasts of hospitalizations in each region, we formulate an allocation algorithm following techniques developed in operations research. We then score forecasts according to how much unmet need their associated allocations would have allowed. We illustrate this scheme with quantile forecasts of COVID-19 hospitalizations in the US at the state level that are recorded in the COVID-19 Forecast Hub, with the goal of determining the allocation of a hypothetical limited resource across the states. The forecast skill ranking given by this allocation scoring rule can vary substantially from the ranking given by the weighted interval score now used by the CDC, especially during surges in hospitalizations such as in late 2021 as the Omicron wave began. We see this as strong evidence that the allocation scoring rule detects forecast value that is missed by traditional accuracy measures and that the general strategy of designing scoring rules directly linked to policy performance is a promising research direction for epidemic forecast evaluation.

@@

Infectious disease forecasting models have emerged as important tools in public health decision making. Predictions of disease dynamics have been used to inform decision-making about a wide variety of measures designed to reduce disease spread and/or mitigate the severity of disease outcomes. For example, estimates of expected onset of flu season were used to inform national vaccination strategies  \citep{igboh2023timing}, and forecasts of ebola dynamics were used to allocate surveillance resources \citep{meltzer2014estimating, rainisch2015regional}. In April of 2022, the Centers for Disease Control and Prevention (CDC) announced the launch of the Center for Forecasting and Outbreak Analytics (CFA) to translate disease forecasts into decision-making \citep{cdc2022cfa}. \cite{bertsimas2021predictionsCOVID} developed tools to inform decision making from infectious disease forecasts, which have been used to to inform allocation of limited medical supplies such as ventilators, ICU capacity planning, and vaccine distribution strategy. Models developed by \cite{fox_real-time_2022} have been used to inform resource and care site planning, as well as community guidelines for masking, traveling, dining and shopping \citep{utnews2022}.  \bwr{Figure out these news site citations!}

@@

@@

In practice, probabilistic infectious disease forecasts have most often been made for observations that emerge from public health surveillance systems and evaluated with standard, ``off-the-shelf'' scoring rules.
For example, seasonal influenza forecasts in the US and dengue forecasts for Peru and Puerto Rico targeted public health surveillance measures of incidence over time and space, and used log-score and mean absolute errors to evaluate forecast skill \citep{mcgowan_collaborative_2019,reich_collaborative_2019,johansson_open_2019}.
Pandemic COVID-19 forecasts of observed cases, hospitalizations and/or deaths in the US and Europe, as reported by municipal, state, or federal surveillance systems, were evaluated using the weighted interval score (WIS, which is an approximation of the continuous ranked probability score, or CRPS), and prediction interval coverage \citep{cramer_evaluation_2022,fox_real-time_2022,sherratt2023predictive}.
Similarly, CRPS was also used to assess probabilistic forecasts of dengue incidence at the district level in Vietnam \citep{colon-gonzalez_probabilistic_2021}.
While some of these scores can be interpreted through the lens of decision theory, and all of the application-specific papers cited above had authors from public health agencies, none of them (including some for which we were co-authors) make explicit connections between forecast evaluation and how a forecast was used in practice.

The motivating application of this work is one where forecasts of standard public health surveillance metrics (1) are used to help determine the allocation of a limited quantity of medical supplies across multiple regions, and (2) are later evaluated based on their recommended allocations.  
The analysis presented is ``synthetic'' in that it does not correspond to an actual analysis that supported decision-making in real-time.
However, the framework described in this paper corresponds to real-world decisions that must be made by public health administrators around the globe, and could be adapted in the future for such real-time situations.  
For example, forecasts for districts in Sierra Leone of bed demand to care for patients with Ebola was the subject of a real-time modeling study in late 2014 and early 2015\cite{camacho2015-ebola-bed}.
And, in 2020, a model developed by an academic research group turned predictions of COVID-19 hospitalizations into estimates of ventilator usage and and shortages. This framework was used by the Hartford HealthCare system in Connecticut ``to align ventilator supply with projected demand at a time where the [COVID-19] pandemic was on the rise''\cite{bertsimas2021predictionsCOVID}. 
These examples illustrate the potential for forecasts to inform decisions about how to allocate limited supplies such as temporary hospital beds, ventilators, personal protective equipment, or other supplies that are known to be effective at reducing transmission or severity of disease.

@@

As described above, there exists both a rich literature on outbreak forecast evaluation and an emerging body of work on turning forecasts into resource allocation decisions. 
However, we have not found literature exploring whether standard forecast accuracy metrics are sufficient for evaluating how well forecasts support allocation decisions.
We hope to begin in this work to fill this gap between the ways in which infectious disease forecasts have been used to support public health policy and the ways in which they have traditionally been evaluated. 
To this end, we define a new forecast scoring rule --- the {\em allocation score} --- that explicitly evaluates forecasts based on how beneficial their implied allocations would turn out to be rather than some standard statisitical measure of their accuracy relative to the surveillance data being forecasted.

@@

While multivariate scoring rules have not seen wide application in infectious disease forecast evalution, it seems that the allocation score would have a similar relationship with them, since while bias penalties in multi-variate scores can be offset by better forecasting of dependence structure, the proportional biases that allocation scoring are insensitive to are not \emph{per se} tolerated by multivariate scoring rules such as the energy score, variogram score, or Dawid-Sebastiani score.

\subsection{Organization}

The remainder of this article is organized as follows. 
In Section \ref{sec:methods}, we review the general framework for developing scoring rules for probabilistic forecasts using the tools of decision theory, develop a novel scoring rule that is motivated by the problem of allocating limited medical supplies, and explore the relationship between the proposed allocation score and existing scoring rules such as CRPS. 
In Section \ref{sec:application} we then illustrate the scoring rule through an application to short-term forecasts of COVID-19 hospital admissions in the US. 
Section \ref{sec:discussion} summarizes our contributions and discusses opportunities for further extensions in future work.

>>>>

Here we pursue one possible tether between multivariate forecasts and policy: the allocation of limited medical resources in response to COVID-19 hospitalizations in various regions so as to minimize expected unmet need. Given probabilistic forecasts of hospitalizations in each region, we formulate an allocation algorithm following techniques developed in operations research. We then score forecasts according to how much unmet need their associated allocations would have allowed. We illustrate this scheme with quantile forecasts of COVID-19 hospitalizations in the US at the state level that are recorded in the COVID-19 Forecast Hub, with the goal of determining the allocation of a hypothetical limited resource across the states. The forecast skill ranking given by this allocation scoring rule can vary substantially from the ranking given by the weighted interval score now used by the CDC, especially during surges in hospitalizations such as in late 2021 as the Omicron wave began. We see this as strong evidence that the allocation scoring rule detects forecast value that is missed by traditional accuracy measures and that the general strategy of designing scoring rules that are directly linked to policy performance is a promising research direction for epidemic forecast evaluation.

@@

Infectious disease forecasting models have emerged as important tools in public health decision making. Predictions of disease dynamics have been used to inform decision-making about a wide variety of measures designed to reduce disease spread and/or mitigate the severity of disease outcomes. For example, estimates of expected onset of flu season were used to inform national vaccination strategies  \citep{igboh2023timing}, and forecasts of ebola dynamics were used to allocate surveillance resources \citep{meltzer2014estimating, rainisch2015regional}. \cite{bertsimas2021predictionsCOVID} developed tools to inform decision making from infectious disease forecasts, which have been used to inform allocation of limited medical supplies such as ventilators, ICU capacity planning, and vaccine distribution strategy. Models developed by \cite{fox_real-time_2022} have been used to inform resource and care site planning, as well as community guidelines for masking, traveling, dining and shopping \citep{utnews2022}. In April of 2022, the Centers for Disease Control and Prevention (CDC) announced the launch of the Center for Forecasting and Outbreak Analytics (CFA) to translate disease forecasts into decision-making \citep{cdc2022cfa}, indicating that this has been identified as an important direction at the highest levels of government public health response. \bwr{Figure out these news site citations!}

@@

% Other important early contributions include Nelson (1972) and White (1966), who are amongst the first to consider the problem of forecast evaluation from the perspective of decision theory and dynamic programming. Nelson, in particular, articulates the concept of the “value of a (probability) fore- cast” to an individual decision-maker as the expected value of his/her payoff using the probability forecasts relative to the payoff expected if unconditional probability estimates are used.

% This has been somewhat rectified during the last decade as the observation that financial series are predictable has led to an attempt to understand the economic significance of this finding. See, for example, Breen, Glosten, and Jagannathan (1989), Leitch and Tanner (1991), McCulloch and Rossi (1990), West, Edison, and Cho (1993), Pesaran and Timmermann (1994, 1995), Satchell and Timmermann (1995), and Skouras (1998).

% Theil (1960), White (1966)

@@

In practice, probabilistic infectious disease forecasts have most often been made for observations that emerge from public health surveillance systems and have typically been evaluated with standard, ``off-the-shelf'' scoring rules.
For example, seasonal influenza forecasts in the US and dengue forecasts for Peru and Puerto Rico targeted public health surveillance measures of incidence over time and space, and used log-score and mean absolute errors to evaluate forecast skill \citep{mcgowan_collaborative_2019,reich_collaborative_2019,johansson_open_2019}.
Pandemic COVID-19 forecasts of observed cases, hospitalizations and/or deaths in the US and Europe, as reported by municipal, state, or federal surveillance systems, were evaluated using the weighted interval score (WIS, which is an approximation of the continuous ranked probability score, or CRPS), and prediction interval coverage \citep{cramer_evaluation_2022,fox_real-time_2022,sherratt2023predictive}.
Similarly, CRPS was also used to assess probabilistic forecasts of dengue incidence at the district level in Vietnam \citep{colon-gonzalez_probabilistic_2021}.
While some of these scores can be interpreted through the lens of decision theory, and all of the application-specific papers cited above had authors from public health agencies, none of them make explicit connections between forecast evaluation and how a forecast was used in practice.

In this work, we hope to begin to fill this gap between the ways in which infectious disease forecasts have been used to support public health policy and the ways in which they have traditionally been evaluated.
For concreteness, we consider a motivating application in which forecasts of public health surveillance signals are used to help determine the allocation of a limited quantity of medical supplies across multiple regions.
We define a new forecast scoring rule --- the {\em allocation score} --- that explicitly evaluates forecasts based on how beneficial their implied allocations would turn out to be.
Unlike the traditional measures of forecast skill such as WIS, CRPS, and log score, the proposed allocation score is rooted in this decision making context and measures the suitability of a forecast as an input to that specific decision making task.

We present an illustrative analysis using the allocation score to evaluate forecasts of hospital admissions in the US leading up to the Omicron wave in winter 2022.
This analysis is ``synthetic'' in that it does not correspond to an actual analysis that supported decision-making in real-time.
However, the framework described in this paper corresponds to real-world decisions that must be made by public health administrators around the globe, and could be adapted in the future for such real-time situations.
For example, forecasts for districts in Sierra Leone of bed demand to care for patients with Ebola was the subject of a real-time modeling study in late 2014 and early 2015 \citep{camacho2015-ebola-bed}.
And, in 2020, a model developed by an academic research group turned predictions of COVID-19 hospitalizations into estimates of ventilator usage and and shortages. This framework was used by the Hartford HealthCare system in Connecticut ``to align ventilator supply with projected demand at a time where the [COVID-19] pandemic was on the rise'' \citep{bertsimas2021predictionsCOVID}.
These examples illustrate the potential for forecasts to inform decisions about how to allocate limited supplies such as temporary hospital beds, ventilators, personal protective equipment, or other supplies that are known to be effective at reducing transmission or severity of disease.
However, we emphasize again that these studies did not take the step of evaluating forecasts based on the quality of the allocation decisions that they supported or could have been used to support.

% As described above, there exists both a rich literature on outbreak forecast evaluation and an emerging body of work on turning forecasts into resource allocation decisions. 
% However, we have not found literature exploring whether standard forecast accuracy metrics are sufficient for evaluating how well forecasts support allocation decisions.

% \subsection{Organization}

The remainder of this article is organized as follows. 
In Section \ref{sec:methods}, we review the general framework for developing scoring rules for probabilistic forecasts using the tools of decision theory, develop a novel scoring rule that is motivated by the problem of allocating limited medical supplies, and explore the relationship between the proposed allocation score and existing scoring rules such as CRPS. 
In Section \ref{sec:application} we illustrate the scoring rule through an application to short-term forecasts of COVID-19 hospital admissions in the US. 
Section \ref{sec:discussion} summarizes our contributions and discusses opportunities for further extensions in future work.

@@

@@

While multivariate scoring rules have not seen wide application in infectious disease forecast evaluation, it seems that the allocation score would have a similar relationship with them, since while bias penalties in multi-variate scores can be offset by better forecasting of dependence structure, the proportional biases that allocation scoring are insensitive to are not \emph{per se} tolerated by multivariate scoring rules such as the energy score, variogram score, or Dawid-Sebastiani score.
